

## Hierarchical Clustering (agglom), using Ward's minimum variance criteria

```{r, echo = FALSE, include=FALSE}
# Load data 
load("../data/df_mds_coords.RData")
load("../data/df_mds_result.RData")
load("../data/df_2023_2025_combined.RData")


### clustering (Using MDS results)
library(clusterSim)
library(cluster)     # For clustering and silhouette
library(factoextra)  # For visualization
library(FactoMineR)  # For HCPC clustering
library(dplyr)       # For data manipulation
library(ggplot2)     # For visualization
library(mclust)      # For ARI
library(here)

```

### Evaluate clustering tendency 

If the value of Hopkins statistic is near 0.5, the data are uniformly distributed. If the value of the Hopkins statistic is far from 0.5, the data are clusterable

```{r}
#MDS results
res_mds <- get_clust_tendency(df_mds_result, n = nrow(df_mds_result)-1, graph = FALSE)
print(res_mds)
```

## Outlier Detection 

```{r}
conVars <- c("Dim1", "Dim2")
IsOutlier <- function(x) {
  outlier <- (x < quantile(x, 0.25, na.rm = TRUE) - 3 * IQR(x, na.rm = TRUE)) |
    (x > quantile(x, 0.75, na.rm = TRUE) + 3 * IQR(x, na.rm = TRUE))
  as.numeric(outlier)
}

# Identify outliers for continuous variables
for (i in names(conVars)) {
  mds_coords[[paste0("Outlier_", i)]] <- IsOutlier(Data.Cluster[[i]])
}

# Create Exclude column
mds_coords$Exclude <- rowSums(mds_coords[, grepl("Outlier", names(mds_coords))], na.rm = TRUE)

# Check for outliers
has_outliers <- any(mds_coords$Exclude != 0)
cat("Does this data set contain outliers?", has_outliers, "\n")
```

## Separate Outliers (if any)

```{r}
if (has_outliers) {
  # Separate outliers
  outliers <- mds_coords[Data.Cluster$Exclude != 0, ]
  cat("The following outliers were removed: \n")
  print(outliers)
  
  # Remove outliers from data sets 
  mds_coords <- mds_coords[mds_coords$Exclude == 0, ]
  df_mds_result <- df_mds_result[mds_coords$Exclude == 0, ]
    
  cat("Cleaned dataset prepared:", nrow(mds_coords), "\n")
  
} else {
  cat("No outliers detected. Proceed with analysis.\n")
}
```

## Test all possible combinations of methods for best silhouette score

```{r, echo=FALSE}
library(clValid)
distmethod = c("euclidean","manhattan")
clmethods = c("kmeans","pam","diana", "hierarchical")

for (i in distmethod) {
      dist = daisy(df_mds_result, metric = i, stand = TRUE)
      distmat = as.matrix(dist)
      intern = clValid(distmat, nClust = 2:10, clMethods = clmethods, method = "complete", validation = "internal", maxitems = 500, metric = i)
print(paste(i))
summary(intern)
}
```

Hierarchical clustering with Manhattan distance results in optimal silhouette scores. 

## Compute Manhattan distance on MDS results 

```{r}
dist_matrix_mds <- dist(df_mds_result, method = "manhattan")
```

# Method 1: Perform hierarchical clustering using Ward's method 

```{r}
hc_ward <- hclust(dist_matrix_mds, method = "ward.D2")

#Determine number of clsuters 

# Computes average silhouette of observations for different values of k
  fviz_nbclust(
    df_mds_result,                   
    FUNcluster = hcut,            # Hierarchical clustering 
    method = "silhouette", 
    diss = dist_matrix_mds,
    k.max = 10,                   # Maximum number of clusters to test
    hc_method = "ward.D2"        # Linkage method 
  )
  
# Elbow Method 
  fviz_nbclust(
    df_mds_result,
    FUNcluster = hcut,
    method = "wss",
    diss = dist_matrix_mds,
    k.max = 10,
    hc_method = "ward.D2"
  )

```

### Dendrogram to inspect number of clusters 

```{r}
# Plot dendrogram
plot(hc_ward, main = "Dendrogram (Ward's Method)", sub = "", xlab = "", cex = 0.6)
```

### Cut dendrogram into k clusters 

Note: manually edit number of clusters based on dendrogram and silhouette widths 

```{r}
clusters <- cutree(hc_ward, k = 2)

# Add clusters to data
mds_coords$cluster <- factor(clusters)

# Plot clusters
ggplot(mds_coords, aes(x = Dim1, y = Dim2, color = cluster)) +
  geom_point(size = 3) +
  labs(title = "Hierarchical Clustering on MDS Coordinates") +
  theme_minimal()
```

### Compute silhouette scores
```{r}
sil_scores <- silhouette(clusters, dist_matrix_mds)
summary(sil_scores)
# Plot silhouette scores
sil_width_plot <- fviz_silhouette(sil_scores)
sil_width_plot
ggsave("../outputs/sil_width_plot.png", plot = sil_width_plot, width = 8, height = 6, dpi = 300)
```

### Evaluate Cluster Stability 

Clusterwise Jaccard subsetting mean: < 0.6 unstable 0.6 - 0.75 low stability 0.75 - 0.85 moderate stablity > 0.85 stable

```{r}
library(fpc)

kbest = 2 #can edit according to number of clusters chosen 
perc = 0.8
s = 987

# Run clusterboot for stability
cboot_HCPC = clusterboot(dist_matrix_mds, clustermethod = hclustCBI, method = "ward.D2", k = kbest, distances = TRUE, bootmethod = "subset", subtuning = nrow(df_mds_result)*perc,B=200, seed= s, count = FALSE)

print(cboot_HCPC)

# Save results
# Extract components
jaccard <- cboot_HCPC$subsetmean
dissolved <- cboot_HCPC$subsetbrd
recovered <- cboot_HCPC$subsetrecover

# Combine into a data frame
cboot_df <- data.frame(
  Cluster = seq_along(jaccard),
  Jaccard = jaccard,
  Dissolved = dissolved,
  Recovered = recovered
)

overall_mean_jaccard_HCPC <- mean(jaccard, na.rm = TRUE)

jaccard_summary_HCPC <- rbind(
  cboot_df,
  data.frame(
    Cluster = "Overall_Mean",
    Jaccard = overall_mean_jaccard_HCPC,
    Dissolved = NA,
    Recovered = NA
  )
)
print(jaccard_summary_HCPC)
```

# Plot Waveforms by cluster 

```{r}
# Merge cluster assignment to waveform data 
dat_clust_plot <- mds_coords %>%
  dplyr::select(c("subject", "cluster")) %>%
  left_join(df_combined, join_by("subject"))
  
component_names <- c(
  `X`="Sagittal  Plane",
  `Y`="Frontal Plane",
  `Z`="Transverse Plane")

# Plot
library(viridis)

plot_WFbyClust_mds_front <- dat_clust_plot %>%
  filter(signal_names == "KNEE_ANGLE") %>%
  filter(signal_components == "Y") %>%
  filter(item %in% (1:60)) %>%
  ggplot(aes(x = item, y = value, color = cluster, fill = cluster)) +  # Map both color and fill
  geom_hline(yintercept = 0, color = "black", linewidth = 0.25) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # stat_summary(
  #   fun.data = "mean_sdl",
  #   fun.args = list(mult = 1),
  #   geom = "ribbon",
  #   alpha = 0.25) +  # Fill will match color
  facet_wrap(signal_names ~ signal_components, 
             scales = "free", 
             labeller = labeller(signal_components = as_labeller(component_names)), 
             ncol = 3) +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank(),
    legend.position = 'right',
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(title = "Mean Waveforms Coloured by Cluster",
       x = "Gait Cycle (%)",
       y = "Mean Waveform Value",
       color = "Cluster",
       fill = "Cluster") +  # Ensure fill is also labeled
  scale_color_viridis(discrete = TRUE, option = "D") +  # Viridis for lines
  scale_fill_viridis(discrete = TRUE, option = "D")

plot_WFbyClust_mds_front
#ggsave("../outputs/HC_WFbyClust_plot_front.png", plot = plot_WFbyClust_mds_front, width = 8, height = 6, dpi = 300)

plot_WFbyClust_mds_front_PP <- dat_clust_plot %>%
  filter(signal_names == "KNEE_ANGLE") %>%
  filter(signal_components == "Y") %>%
  filter(item %in% (1:60)) %>%
  ggplot(aes(x = item, y = value, color = cluster, fill = cluster)) +  # Map both color and fill
  geom_hline(yintercept = 0, color = "black", linewidth = 0.25) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # stat_summary(
  #   fun.data = "mean_sdl",
  #   fun.args = list(mult = 1),
  #   geom = "ribbon",
  #   alpha = 0.25) +  # Fill will match color
  facet_wrap(signal_names ~ signal_components, 
             scales = "free", 
             labeller = labeller(signal_components = as_labeller(component_names)), 
             ncol = 3) +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank(),
    text = element_text(size = 20),
    legend.position = 'right'
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(title = "Mean Waveforms Coloured by Cluster",
       x = "Gait Cycle (%)",
       y = "Mean Waveform Value",
       color = "Cluster",
       fill = "Cluster") +  # Ensure fill is also labeled
  scale_color_viridis(discrete = TRUE, option = "D") +  # Viridis for lines
  scale_fill_viridis(discrete = TRUE, option = "D")

plot_WFbyClust_mds_front_PP
#ggsave("../outputs/HC_WFbyClust_plot_front_PP.png", plot = plot_WFbyClust_mds_front_PP, width = 8, height = 6, dpi = 300)


plot_WFbyClust_mds_sag <- dat_clust_plot %>%
  filter(signal_names == "KNEE_ANGLE") %>%
  filter(signal_components == "X") %>%
  ggplot(aes(x = item, y = value, color = cluster, fill = cluster)) +  # Map both color and fill
  geom_hline(yintercept = 0, color = "black", linewidth = 0.25) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # stat_summary(
  #   fun.data = "mean_sdl",
  #   fun.args = list(mult = 1),
  #   geom = "ribbon",
  #   alpha = 0.25) +  # Fill will match color
  facet_wrap(signal_names ~ signal_components, 
             scales = "free", 
             labeller = labeller(signal_components = as_labeller(component_names)), 
             ncol = 3) +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank(),
    legend.position = 'right',
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(title = "Mean Waveforms Coloured by Cluster",
       x = "Gait Cycle (%)",
       y = "Mean Waveform Value",
       color = "Cluster",
       fill = "Cluster") +  # Ensure fill is also labeled
  scale_color_viridis(discrete = TRUE, option = "D") +  # Viridis for lines
  scale_fill_viridis(discrete = TRUE, option = "D")

plot_WFbyClust_mds_sag
#ggsave("../outputs/HC_WFbyClust_plot_sag.png", plot = plot_WFbyClust_mds_sag, width = 8, height = 6, dpi = 300)

plot_WFbyClust_mds_sag_PP <- dat_clust_plot %>%
  filter(signal_names == "KNEE_ANGLE") %>%
  filter(signal_components == "X") %>%
  ggplot(aes(x = item, y = value, color = cluster, fill = cluster)) +  # Map both color and fill
  geom_hline(yintercept = 0, color = "black", linewidth = 0.25) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # stat_summary(
  #   fun.data = "mean_sdl",
  #   fun.args = list(mult = 1),
  #   geom = "ribbon",
  #   alpha = 0.25) +  # Fill will match color
  facet_wrap(signal_names ~ signal_components, 
             scales = "free", 
             labeller = labeller(signal_components = as_labeller(component_names)), 
             ncol = 3) +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank(),
    text = element_text(size = 20),
    legend.position = 'right'
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(title = "Mean Waveforms Coloured by Cluster",
       x = "Gait Cycle (%)",
       y = "Mean Waveform Value",
       color = "Cluster",
       fill = "Cluster") +  # Ensure fill is also labeled
  scale_color_viridis(discrete = TRUE, option = "D") +  # Viridis for lines
  scale_fill_viridis(discrete = TRUE, option = "D")

plot_WFbyClust_mds_sag_PP
#ggsave("../outputs/HC_WFbyClust_plot_sag_PP.png", plot = plot_WFbyClust_mds_sag_PP, width = 8, height = 6, dpi = 300)
```
# Cluster Characteristics

```{r, echo = FALSE}
library(dplyr)
library(tableone)
library(pander)
load("../data/mds_input_data.RData")

# Edit based on PCA results (only PCs with >= 0.8 Culm Var proportion are included)
variable_list <- c("age", "PC1_front", "PC1_sag", "PC2_sag", "PC3_sag", "speed", "sex", "severity_contra", "oa_location", "signal_side", "kl_contra")
factor_variables <- c("sex", "severity_contra", "kl_contra", "oa_location", "signal_side")
continuous_vars <- c("age", "PC1_front", "PC1_sag", "PC2_sag", "PC3_sag", "speed")
 
dat_clust_tables <- mds_input_data %>%
  left_join(mds_coords[, c("subject", "cluster")], join_by("subject"))
dat_clust_tables <- dat_clust_tables %>%
  left_join(
    df_combined %>%
      group_by(subject) %>%
      summarise(
        sex = first(sex),
        oa_location = first(oa_location),
        severity_contra = first(severity_contra),
        kl_contra = first(kl_contra)
      ) %>%
      dplyr::select(subject, sex, oa_location, severity_contra, kl_contra), 
    by = "subject"
    
    )

print("Overall Sample Characteristics")
tab0 <- CreateTableOne(vars = variable_list,, data = dat_clust_tables, factorVars = factor_variables, test = FALSE)
print(tab0,  formatOptions = list(big.mark = ","))
```

```{r, echo = FALSE, include = FALSE}
# tab0_matrix <- print(tab0,
#   includeNA = TRUE,
#   showAllLevels = TRUE
# )
# 
# write.csv(tab0_matrix, file = "../outputs/HCPC_sample_summary.csv")
```

```{r}
## cluster tables
print("Cluster Characteristics")
tab1 <- CreateTableOne(vars = variable_list, strata = "cluster" , data = dat_clust_tables, factorVars = factor_variables, test = FALSE)
print(tab1)
```

```{r, include = FALSE, echo = FALSE}
# tab1_matrix <- print(tab1,
#   includeNA = TRUE,
#   showAllLevels = TRUE
# )
# 
# #convert to .csv for report 
# write.csv(tab1_matrix, file = "../outputs/HCPC_cluster_summaries.csv")
```

```{r}
# Check how many subjects are missing KL grades
print("Check how many subjects are missing KL grades")
print(sum(is.na(dat_clust_tables$kl_contra)))
dat_clust_tables %>%
  filter(is.na(kl_contra)) %>%
  distinct(subject, cluster) %>%
  group_by(cluster) %>%
  summarise(n_missing = n())
```

# Statistical tests on Clusters

## Continuous Variables 

```{r}
library(broom)

test_continuous_var <- function(data, variable, cluster_col = "cluster", file_prefix = NULL) {
  var_sym <- rlang::sym(variable)
  cluster_sym <- rlang::sym(cluster_col)
  
  # Normality check
  shapiro_test <- data %>%
    group_by(!!cluster_sym) %>%
    filter(n() >= 3) %>% # exclude clusters with less than 3 observations 
    summarise(p_value = shapiro.test(!!var_sym)$p.value, .groups = "drop")
  print(shapiro_test)
  
  # Check excluded clusters
  excluded <- data %>%
    group_by(!!cluster_sym) %>%
    filter(n() < 3) %>%
    summarise(n = n(), .groups = "drop")
  print(excluded)
  
  # Determine which test to run
  if (any(shapiro_test$p_value < 0.05)) {
    test_result <- kruskal.test(reformulate(cluster_col, response = variable), data = data)
    test_type <- "Kruskal-Wallis"
    test_p <- test_result$p.value
  } else {
    test_result <- aov(reformulate(cluster_col, response = variable), data = data)
    test_type <- "ANOVA"
    test_summary <- summary(test_result)
    test_p <- test_summary[[1]][["Pr(>F)"]][1]
  }
  
  cat(paste0("\n", test_type, " test for ", variable, ":\n"))
  print(test_result)
  
  # Save main test result (optional -- if signif, pairwise tests will be saved)
  if (!is.null(file_prefix)) {
    tidy_test <- tidy(test_result)
    write.csv(tidy_test, file = paste0("../outputs/", file_prefix, "_", test_type, ".csv"), row.names = FALSE)
  }
  
  # Plot boxplot
  boxplot(as.formula(paste(variable, "~", cluster_col)), data = data,
          main = paste(variable, "Across Clusters"), xlab = "Cluster")
  
  # Pairwise test if significant
  if (test_p < 0.05 && length(unique(data[[cluster_col]])) > 2) {
    if (inherits(test_result, "aov")) {
      tukey_result <- TukeyHSD(test_result)
      print(tukey_result)
      pairwise_df <- as.data.frame(tukey_result[[cluster_col]])
      pairwise_df$Comparison <- rownames(pairwise_df)
      pairwise_df <- pairwise_df[, c("Comparison", names(pairwise_df)[1:(ncol(pairwise_df)-1)])]
      method <- "TukeyHSD"
    } else {
      wilcox_result <- pairwise.wilcox.test(data[[variable]], data[[cluster_col]], p.adjust.method = "bonferroni")
      print(wilcox_result)
      pairwise_df <- tidy(wilcox_result)
      method <- "Wilcoxon"
    }
    
    # Save pairwise result
    if (!is.null(file_prefix)) {
      write.csv(pairwise_df, file = paste0("../outputs/", file_prefix, "_", method, ".csv"), row.names = FALSE)
    }
  }
}
```

```{r, warning = FALSE}
# Run test 
# Ensure cluster is a factor
dat_clust_tables <- dat_clust_tables %>% mutate(cluster = as.factor(cluster))

# Edit for PCs included 
print("AGE")
test_continuous_var(dat_clust_tables, "age", file_prefix = "k2_HCPC_Age")
print("SPEED")
test_continuous_var(dat_clust_tables, "speed", file_prefix = "k2_HCPC_Speed")
print("PC1 FRONTAL PLANE")
test_continuous_var(dat_clust_tables, "PC1_front", file_prefix = "k2_HCPC_PC1Front")
print("PC1 SAGITTAL PLANE")
test_continuous_var(dat_clust_tables, "PC1_sag", file_prefix = "k2_HCPC_PC1Sag")
print("PC2 SAGITTAL PLANE")
test_continuous_var(dat_clust_tables, "PC2_sag", file_prefix = "k2_HCPC_PC2Sag")
print("PC3 SAGITTAL PLANE")
test_continuous_var(dat_clust_tables, "PC3_sag", file_prefix = "k2_HCPC_PC3Sag")
```

# Categorical Variables

```{r}
test_categorical <- function(var, cluster_col = "final_clusters") {
  # Create contingency table of unique subjects
  unique_data <- dat_clust_tables %>%
    distinct(subject, !!sym(cluster_col), !!sym(var))
  
  # Extract vectors for table
  var_vector <- pull(unique_data, !!sym(var))
  cluster_vector <- pull(unique_data, !!sym(cluster_col))
  
  contingency_table <- table(var_vector, cluster_vector)
  
  # Print the table
  cat(sprintf("\nContingency table for %s (unique subjects):\n", var))
  print(contingency_table)
  # Save contingency table as CSV
  contingency_df <- as.data.frame.matrix(contingency_table)
  contingency_df$Category <- rownames(contingency_df)
  write.csv(contingency_df, file = sprintf("../outputs/HCPC_%s_%s_contingency_table.csv", cluster_col, var), row.names = FALSE)
  
  # Chi-squared test
  chi_test <- chisq.test(contingency_table)
  cat(sprintf("\nChi-squared test for %s:\n", var))
  print(chi_test)
  chi_df <- broom::tidy(chi_test)
  write.csv(chi_df, file = sprintf("../outputs/HCPC_%s_%s_chi_squared.csv", cluster_col, var), row.names = FALSE)
  
  # Check expected counts
  expected <- chi_test$expected
  if (any(expected < 5)) {
    cat(sprintf("\nWarning: Some expected counts < 5 for %s, performing Fisher's exact test.\n", var))
    fisher_test <- fisher.test(contingency_table, simulate.p.value = TRUE)
    cat(sprintf("Fisher's exact test for %s:\n", var))
    print(fisher_test)
    fisher_df <- broom::tidy(fisher_test)
    write.csv(fisher_df, file = sprintf("../outputs/HCPC_%s_%s_fisher_exact.csv", cluster_col, var), row.names = FALSE)
  }
}
```

```{r, warning = FALSE}
# Run tests for each categorical variable
lapply(factor_variables, test_categorical, cluster_col = "cluster")
```


## Method 2: Evaluate with Silhouette and adjusted Rand Index (ARI) 

Method from Young-Shand et al. 2023:

"The cluster model of the full data set was compared with 100 subsets of the full data set (fraction of 0.8) for k = 2:10, and the mean ARI was reported. The selected clustering  solution (k clusters) with the greatest mean s and ARI was chosen."

```{r, echo = FALSE}
library(dplyr)
library(cluster)  # For silhouette
library(mclust)   # For adjustedRandIndex
library(stats)    # For hclust
```

```{r}
# Hierarchical clustering function
hclust_function <- function(data, k) {
  dist_matrix <- dist(data, method = "manhattan")
  hc <- hclust(dist_matrix, method = "ward.D2")
  clusters <- cutree(hc, k = k)
  return(clusters)
}

# Parameters
n_iterations <- 100  # Number of subsamples
fraction <- 0.8      # 80% of data
k_range <- 2:10      # Test k from 2 to 10
n_samples <- round(nrow(df_mds_result) * fraction)  # Number of rows to sample

# Store results
ari_results <- matrix(NA, nrow = n_iterations, ncol = length(k_range))
sil_results <- matrix(NA, nrow = n_iterations, ncol = length(k_range))
colnames(ari_results) <- paste("k =", k_range)
mean_sil <- numeric(length(k_range))  # Mean silhouette for full data

set.seed(123)  # For reproducibility

# Cluster full dataset for each k
full_clusters <- lapply(k_range, function(k) hclust_function(df_mds_result, k))

# Subsampling and ARI calculation
for (i in 1:n_iterations) {
  # Subsample 80% of the data
  sample_idx <- sample(1:nrow(df_mds_result), size = n_samples, replace = FALSE)
  sampled_data <- df_mds_result[sample_idx, ]
  
  # Distance Matrix 
  diss_mat <- dist(sampled_data, method = "manhattan")
  
  # Cluster subsample for each k and compute ARI
  for (j in seq_along(k_range)) {
    k <- k_range[j]
    subsample_clusters <- hclust_function(sampled_data, k)
    full_clusters_sub <- full_clusters[[j]][sample_idx]  # Subset full clusters to match
    ari_results[i, j] <- adjustedRandIndex(full_clusters_sub, subsample_clusters)
    sil <- silhouette(subsample_clusters, diss_mat)
    sil_results[i,j] <- mean(sil[, 3], na.rm = TRUE)
  }
}

# Compute mean sil and ARI for each k
mean_sil <- colMeans(sil_results, na.rm = TRUE)
names(mean_sil) <- paste("k =", k_range)

mean_ari <- colMeans(ari_results, na.rm = TRUE)
names(mean_ari) <- paste("k =", k_range)

# Print results
cat("\nMean Adjusted Rand Index (Stability) for each k:\n")
print(round(mean_ari, 3))

cat("\nMean Silhouette Width for each k (Full Dataset):\n")
print(round(mean_sil, 3))

# Save results to csv 
subsampling_results <- data.frame(
  k = k_range,
  Mean_ARI = round(mean_ari, 3),
  Mean_Silhouette = round(mean_sil, 3)
)

write.csv(subsampling_results, file = "../outputs/subsampling_results.csv", row.names = FALSE)
```

### Select optimal k (highest mean ARI and silhouette)

```{r}
combined_score <- mean_ari + mean_sil  # Simple sum (could adjust weighting)
optimal_k_idx <- which.max(combined_score)
optimal_k <- k_range[optimal_k_idx]
cat("\nSelected k with highest combined mean ARI and silhouette:", optimal_k, "\n")
cat("Mean ARI for selected k:", round(mean_ari[optimal_k_idx], 3), "\n")
cat("Mean Silhouette for selected k:", round(mean_sil[optimal_k_idx], 3), "\n")
```

### Final clustering with optimal k

```{r}
final_clusters <- hclust_function(df_mds_result, optimal_k)

# Add final clusters to data
mds_coords$final_clusters <- factor(final_clusters)
save(mds_coords, file="../data/df_HC_mds_coords.RData")

## Create cluster plot based on HCPC
HC_MDS_plot <- fviz_cluster(
  list(data = mds_coords[, c("Dim1", "Dim2")], cluster = mds_coords$final_clusters),
  repel = TRUE,
  show.clust.cent = FALSE,
  show.legend.text = FALSE,
  pointsize = 1.8,
  geom = "point",
  ggtheme = theme_minimal() +
    theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5)),
  main = "Hierarchical Clustering on MDS Coordinates"
) +
  theme_bw()
HC_MDS_plot
ggsave("../outputs/HC_MDS_plot.png", plot = HC_MDS_plot, width = 8, height = 6, dpi = 300)

HC_MDS_plot_v2 <- ggplot(mds_coords, aes(x = Dim1, y = Dim2, color = final_clusters)) +
  geom_point(size = 3) +
  labs(title = "Hierarchical Clustering on MDS Coordinates") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
HC_MDS_plot_v2
ggsave("../outputs/HC_MDS_plot_v2.png", plot = HC_MDS_plot_v2, width = 8, height = 6, dpi = 300)
```

# Plot Waveforms by cluster 

```{r}
# Merge cluster assignment to waveform data 
dat_clust_plot <- mds_coords %>%
  dplyr::select(c("subject", "final_clusters")) %>%
  left_join(df_combined, join_by("subject"))
  
component_names <- c(
  `X`="Sagittal  Plane",
  `Y`="Frontal Plane",
  `Z`="Transverse Plane")

# Plot
library(viridis)

plot_WFbyClust_mds_front <- dat_clust_plot %>%
  filter(signal_names == "KNEE_ANGLE") %>%
  filter(signal_components == "Y") %>%
  filter(item %in% (1:60)) %>%
  ggplot(aes(x = item, y = value, color = final_clusters, fill = final_clusters)) +  # Map both color and fill
  geom_hline(yintercept = 0, color = "black", linewidth = 0.25) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # stat_summary(
  #   fun.data = "mean_sdl",
  #   fun.args = list(mult = 1),
  #   geom = "ribbon",
  #   alpha = 0.25) +  # Fill will match color
  facet_wrap(signal_names ~ signal_components, 
             scales = "free", 
             labeller = labeller(signal_components = as_labeller(component_names)), 
             ncol = 3) +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank(),
    legend.position = 'right',
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(title = "Mean Waveforms Coloured by Cluster",
       x = "Gait Cycle (%)",
       y = "Mean Waveform Value",
       color = "Cluster",
       fill = "Cluster") +  # Ensure fill is also labeled
  scale_color_viridis(discrete = TRUE, option = "D") +  # Viridis for lines
  scale_fill_viridis(discrete = TRUE, option = "D")

plot_WFbyClust_mds_front
ggsave("../outputs/HC_WFbyClust_plot_front.png", plot = plot_WFbyClust_mds_front, width = 8, height = 6, dpi = 300)

plot_WFbyClust_mds_front_PP <- dat_clust_plot %>%
  filter(signal_names == "KNEE_ANGLE") %>%
  filter(signal_components == "Y") %>%
  filter(item %in% (1:60)) %>%
  ggplot(aes(x = item, y = value, color = final_clusters, fill = final_clusters)) +  # Map both color and fill
  geom_hline(yintercept = 0, color = "black", linewidth = 0.25) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # stat_summary(
  #   fun.data = "mean_sdl",
  #   fun.args = list(mult = 1),
  #   geom = "ribbon",
  #   alpha = 0.25) +  # Fill will match color
  facet_wrap(signal_names ~ signal_components, 
             scales = "free", 
             labeller = labeller(signal_components = as_labeller(component_names)), 
             ncol = 3) +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank(),
    text = element_text(size = 20),
    legend.position = 'right'
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(title = "Mean Waveforms Coloured by Cluster",
       x = "Gait Cycle (%)",
       y = "Mean Waveform Value",
       color = "Cluster",
       fill = "Cluster") +  # Ensure fill is also labeled
  scale_color_viridis(discrete = TRUE, option = "D") +  # Viridis for lines
  scale_fill_viridis(discrete = TRUE, option = "D")

plot_WFbyClust_mds_front_PP
ggsave("../outputs/HC_WFbyClust_plot_front_PP.png", plot = plot_WFbyClust_mds_front_PP, width = 8, height = 6, dpi = 300)


plot_WFbyClust_mds_sag <- dat_clust_plot %>%
  filter(signal_names == "KNEE_ANGLE") %>%
  filter(signal_components == "X") %>%
  ggplot(aes(x = item, y = value, color = final_clusters, fill = final_clusters)) +  # Map both color and fill
  geom_hline(yintercept = 0, color = "black", linewidth = 0.25) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # stat_summary(
  #   fun.data = "mean_sdl",
  #   fun.args = list(mult = 1),
  #   geom = "ribbon",
  #   alpha = 0.25) +  # Fill will match color
  facet_wrap(signal_names ~ signal_components, 
             scales = "free", 
             labeller = labeller(signal_components = as_labeller(component_names)), 
             ncol = 3) +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank(),
    legend.position = 'right',
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(title = "Mean Waveforms Coloured by Cluster",
       x = "Gait Cycle (%)",
       y = "Mean Waveform Value",
       color = "Cluster",
       fill = "Cluster") +  # Ensure fill is also labeled
  scale_color_viridis(discrete = TRUE, option = "D") +  # Viridis for lines
  scale_fill_viridis(discrete = TRUE, option = "D")

plot_WFbyClust_mds_sag
ggsave("../outputs/HC_WFbyClust_plot_sag.png", plot = plot_WFbyClust_mds_sag, width = 8, height = 6, dpi = 300)

plot_WFbyClust_mds_sag_PP <- dat_clust_plot %>%
  filter(signal_names == "KNEE_ANGLE") %>%
  filter(signal_components == "X") %>%
  ggplot(aes(x = item, y = value, color = final_clusters, fill = final_clusters)) +  # Map both color and fill
  geom_hline(yintercept = 0, color = "black", linewidth = 0.25) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # stat_summary(
  #   fun.data = "mean_sdl",
  #   fun.args = list(mult = 1),
  #   geom = "ribbon",
  #   alpha = 0.25) +  # Fill will match color
  facet_wrap(signal_names ~ signal_components, 
             scales = "free", 
             labeller = labeller(signal_components = as_labeller(component_names)), 
             ncol = 3) +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank(),
    text = element_text(size = 20),
    legend.position = 'right'
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(title = "Mean Waveforms Coloured by Cluster",
       x = "Gait Cycle (%)",
       y = "Mean Waveform Value",
       color = "Cluster",
       fill = "Cluster") +  # Ensure fill is also labeled
  scale_color_viridis(discrete = TRUE, option = "D") +  # Viridis for lines
  scale_fill_viridis(discrete = TRUE, option = "D")

plot_WFbyClust_mds_sag_PP
ggsave("../outputs/HC_WFbyClust_plot_sag_PP.png", plot = plot_WFbyClust_mds_sag_PP, width = 8, height = 6, dpi = 300)
```


## Evaluate Cluster Stability 

```{r}
library(fpc)

kbest = optimal_k #can edit according to number of clusters chosen 
perc = 0.8
s = 987

# Run clusterboot for stability
cboot_HCPC = clusterboot(dist_matrix_mds, clustermethod = hclustCBI, method = "ward.D2", k = kbest, distances = TRUE, bootmethod = "subset", subtuning = nrow(df_mds_result)*perc,B=200, seed= s, count = FALSE)

print(cboot_HCPC)

# Extract components
jaccard <- cboot_HCPC$subsetmean
dissolved <- cboot_HCPC$subsetbrd
recovered <- cboot_HCPC$subsetrecover

# Combine into a data frame
cboot_df <- data.frame(
  Cluster = seq_along(jaccard),
  Jaccard = jaccard,
  Dissolved = dissolved,
  Recovered = recovered
)

overall_mean_jaccard_HCPC <- mean(jaccard, na.rm = TRUE)

jaccard_summary_HCPC <- rbind(
  cboot_df,
  data.frame(
    Cluster = "Overall_Mean",
    Jaccard = overall_mean_jaccard_HCPC,
    Dissolved = NA,
    Recovered = NA
  )
)
print(jaccard_summary_HCPC)
```

```{r, echo = FALSE}
print("Silhouette Plot for Final Clusters")
sil_HCPC <- silhouette(final_clusters, dist_matrix_mds)
summary(sil_HCPC)
fviz_silhouette(sil_HCPC)
```

# Cluster Characteristics

```{r, echo = FALSE}
library(dplyr)
library(tableone)
library(pander)
load("../data/mds_input_data.RData")

# Edit based on PCA results (only PCs with >= 0.8 Culm Var proportion are included)
variable_list <- c("age", "PC1_front", "PC1_sag", "PC2_sag", "PC3_sag", "speed", "sex", "severity_contra", "oa_location", "signal_side", "kl_contra")
factor_variables <- c("sex", "severity_contra", "kl_contra", "oa_location", "signal_side")
continuous_vars <- c("age", "PC1_front", "PC1_sag", "PC2_sag", "PC3_sag", "speed")
 
dat_clust_tables <- mds_input_data %>%
  left_join(mds_coords[, c("subject", "final_clusters")], join_by("subject"))
dat_clust_tables <- dat_clust_tables %>%
  left_join(
    df_combined %>%
      group_by(subject) %>%
      summarise(
        sex = first(sex),
        oa_location = first(oa_location),
        severity_contra = first(severity_contra),
        kl_contra = first(kl_contra)
      ) %>%
      dplyr::select(subject, sex, oa_location, severity_contra, kl_contra), 
    by = "subject"
    
    )

print("Overall Sample Characteristics")
tab0 <- CreateTableOne(vars = variable_list,, data = dat_clust_tables, factorVars = factor_variables, test = FALSE)
print(tab0,  formatOptions = list(big.mark = ","))
```

```{r, echo = FALSE, include = FALSE}
tab0_matrix <- print(tab0,
  includeNA = TRUE,
  showAllLevels = TRUE
)

write.csv(tab0_matrix, file = "../outputs/HCPC_sample_summary.csv")
```

```{r}
## cluster tables
print("Cluster Characteristics")
tab1 <- CreateTableOne(vars = variable_list, strata = "final_clusters" , data = dat_clust_tables, factorVars = factor_variables, test = FALSE)
print(tab1)
```

```{r, include = FALSE, echo = FALSE}
tab1_matrix <- print(tab1,
  includeNA = TRUE,
  showAllLevels = TRUE
)


#convert to .csv for report 
write.csv(tab1_matrix, file = "../outputs/HCPC_cluster_summaries.csv")
```

```{r}
# Check how many subjects are missing KL grades
print("Check how many subjects are missing KL grades")
print(sum(is.na(dat_clust_tables$kl_contra)))
dat_clust_tables %>%
  filter(is.na(kl_contra)) %>%
  distinct(subject, final_clusters) %>%
  group_by(final_clusters) %>%
  summarise(n_missing = n())
```

# Statistical tests on Clusters

## Continuous Variables 

```{r}
# Ensure cluster is a factor
dat_clust_tables <- dat_clust_tables %>% mutate(cluster = as.factor(final_clusters))

# Edit for PCs included 
test_continuous_var(dat_clust_tables, "age", file_prefix = "HCPC_Age", cluster_col = "final_clusters")
test_continuous_var(dat_clust_tables, "speed", file_prefix = "HCPC_Speed", cluster_col = "final_clusters")
test_continuous_var(dat_clust_tables, "PC1_front", file_prefix = "HCPC_PC1Front", cluster_col = "final_clusters")
test_continuous_var(dat_clust_tables, "PC1_sag", file_prefix = "HCPC_PC1Sag", cluster_col = "final_clusters")
test_continuous_var(dat_clust_tables, "PC2_sag", file_prefix = "HCPC_PC2Sag", cluster_col = "final_clusters")
test_continuous_var(dat_clust_tables, "PC3_sag", file_prefix = "HCPC_PC3Sag", cluster_col = "final_clusters")
```

# Categorical Variables

```{r}
# Run tests for each categorical variable
lapply(factor_variables, test_categorical, cluster_col = "final_clusters")
```

# Optional: Plot with healthy (asymptomatic) controls

```{r}
library(ggplot2)
library(dplyr)
library(RColorBrewer)

load("../data/df_RC_MWF_merged_2023.RData")

# Subjects where both legs are "Asymptomatic" (asymptomatic controls from df_MWF_2023)
asymp_controls <- df_2023 %>%
  filter(item %in% 1:60) %>%  #stance phase filter
  group_by(subject) %>%
  summarise(all_asymp = all(severity == "Asymptomatic")) %>%  # Check if all rows (both legs) are Asymptomatic
  filter(all_asymp) %>%
  pull(subject)  # Get subject IDs

# Extract data
healthy_df <- df_2023 %>%
  filter(subject %in% asymp_controls, 
         item %in% 1:60) %>%  # Stance phase
  mutate(cluster = "Healthy",   # Assign "Healthy" as the cluster label
         group = "Healthy")     # Unified group column

# Prepare clustered data with group column
clust_df <- dat_clust_plot %>%
  filter(item %in% 1:60) %>%  # Ensure same filter
  mutate(group = as.character(final_clusters))  # Convert cluster to character for consistency

#Combine into new data frame
dat_clust_healthy <- bind_rows(clust_df, healthy_df)

# check output 
print(paste("Unique subjects in clustered data:", n_distinct(clust_df$subject)))
print(paste("Unique subjects in healthy controls:", n_distinct(healthy_df$subject)))
print(table(dat_clust_healthy$group))

##Plot Powerpoint version 

plot_WFbyClust_with_controls_PP <- dat_clust_healthy %>%
  filter(signal_names == "KNEE_ANGLE") %>%
  filter(signal_components != "Z") %>%
  ggplot(aes(x = item, y = value, color = group, fill = group)) +  # Use group instead of cluster
  geom_hline(yintercept = 0, color = "black", linewidth = 0.25) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # Uncomment if you want ribbons
  # stat_summary(
  #   fun.data = "mean_sdl",
  #   fun.args = list(mult = 1),
  #   geom = "ribbon",
  #   alpha = 0.25
  # ) +
  facet_wrap(signal_names ~ signal_components, 
             scales = "free", 
             labeller = labeller(signal_components = as_labeller(component_names)), 
             ncol = 3) +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank(),
    text = element_text(size = 20),
    legend.position = 'right',
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(title = "Mean Waveforms by Cluster with\nHealthy Comparison",
       x = "Gait Cycle (%)",
       y = "Mean Waveform Value",
       color = "Cluster",
       fill = "Cluster") +
  scale_color_viridis(discrete = TRUE, option = "D") +  # Viridis for lines
  scale_fill_viridis(discrete = TRUE, option = "D")

# Display plot
print(plot_WFbyClust_with_controls_PP)

##Plot report version 
plot_WFbyClust_with_controls <- dat_clust_healthy %>%
  filter(signal_names == "KNEE_ANGLE") %>%
  filter(signal_components != "Z") %>%
  filter(item %in% 1:60) %>%
  ggplot(aes(x = item, y = value, color = group, fill = group)) +  # Use group instead of cluster
  geom_hline(yintercept = 0, color = "black", linewidth = 0.25) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # Uncomment if you want ribbons
  # stat_summary(
  #   fun.data = "mean_sdl",
  #   fun.args = list(mult = 1),
  #   geom = "ribbon",
  #   alpha = 0.25
  # ) +
  facet_wrap(signal_names ~ signal_components, 
             scales = "free", 
             labeller = labeller(signal_components = as_labeller(component_names)), 
             ncol = 3) +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.background = element_blank(),
    text = element_text(size = 12),
    legend.position = 'right',
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(title = "Mean Waveforms by Cluster with Healthy Comparison",
       x = "Gait Cycle (%)",
       y = "Mean Waveform Value",
       color = "Cluster",
       fill = "Cluster") +
  scale_color_viridis(discrete = TRUE, option = "D") +  # Viridis for lines
  scale_fill_viridis(discrete = TRUE, option = "D")

# Display plot
print(plot_WFbyClust_with_controls)
ggsave("../outputs/HC_WFbyClust_with_controls_plot.png", plot = plot_WFbyClust_with_controls, width = 8, height = 6, dpi = 300)
```

